# Module 2.1: Loading Data - ICARDA Barley Example

## Introduction to the Dataset

In this module, we'll learn how to load typical phenotypic data into R. We'll use a real-world example: data from a study on **275 barley accessions conducted at ICARDA in 2019**. This dataset contains various measurements related to agronomic traits, grain quality, and morphological characteristics.

**Why this dataset?** It's representative of the kind of multi-trait data breeders work with. It allows us to practice loading, inspecting, and performing basic summaries on realistic data. This data comes from ICARDA's valuable work in crop improvement for dry areas.

**Column Descriptions (Partial List - full list would be in a data dictionary):** \* `Taxa`: The identifier for each barley accession (genotype). \* `Area`: Grain area (e.g., mm²). \* `B_glucan`: Beta-glucan content (%), a quality trait. \* `DTH`: Days to Heading (days), an agronomic trait. \* `Fe`: Iron content in grain (ppm), a nutritional trait. \* `FLA`: Flag Leaf Area (cm²). \* `GY`: Grain Yield (e.g., t/ha or kg/plot - units should always be known!). \* `PH`: Plant Height (cm). \* `Protein`: Grain protein content (%). \* `TKW`: Thousand Kernel Weight (grams). \* `Zn`: Zinc content in grain (ppm). \* *(And many others related to grain morphology and plant characteristics...)*

Our goal is to load this data (which is typically stored in a file like a CSV or Excel sheet) into an R data frame so we can start analyzing it.

## Setting Up: Libraries and File Path

First, we need to load the R packages that help us read data. Even though we have previously installed and loaded all packages we will need, in case you are only focusing on reading data, the `readr` package (part of `tidyverse`) is excellent for reading text files like CSVs.

Remember our RStudio Project setup! We will assume the data file is saved in the `data/` subfolder of our project.

```{r setup, include=FALSE, echo=FALSE}
library(here)
```

```{r libs}
# Load the necessary libraries
# 'tidyverse' includes 'readr' (for read_csv) and 'dplyr' (for glimpse, etc.)
library(tidyverse) 
```

## Reading the CSV File

Let's say our barley data is stored in a CSV file named `icarda_barley_2019_pheno.csv`.

```{r csv-run, include=FALSE}
barley_pheno_data <- read_csv(here("data","icarda_barley_2019_pheno.csv"))
```

```{r csv, eval=FALSE}
# Define the path to our data file (relative to the project root)
barley_data_file_path <- "data/icarda_barley_2019_pheno.csv"

# Use read_csv() from the readr package to load the data
barley_pheno_data <- read_csv(barley_data_file_path)
```

## First Look: Inspecting the Loaded Data

It's **CRUCIAL** to always inspect your data immediately after loading it to make sure it looks correct.

1.  **`head()`**: Shows the first few rows (default is 6).

2.  **`dim()`**: Shows the dimensions (number of rows, number of columns).

3.  **`glimpse()`** (from `dplyr`): A great way to see column names, their data types, and the first few values. Better than `str()` for tibbles.

4.  **`summary()`**: Provides basic summary statistics for each column (Min, Max, Mean, Median, Quartiles for numeric; counts for character/factor).

## Understanding Data Types in Our Barley Data

When `glimpse()` runs, you'll see types like: \* `Taxa`: Should be `<chr>` (character) as it's an identifier. \* `Area`, `B_glucan`, `DTH`, `GY`, `PH`, etc.: Should mostly be `<dbl>` (double-precision numeric) as they are measurements.

If `read_csv` misinterprets a numeric column as character (e.g., if there's a text entry like "missing" in a numeric column), you'll need to clean that data or specify column types during import using the `col_types` argument in `read_csv()`.

```{r glimpse}
glimpse(barley_pheno_data)
```

## Quick Summary of a Specific Trait

Let's say we are interested in Grain Yield (`GY`).

```{r summary}
# Access the GY column
yield_values <- barley_pheno_data$GY

# Calculate some basic statistics
# na.rm=TRUE ignores missing values in calculation
mean_yield <- mean(yield_values, na.rm = TRUE)
min_yield <- min(yield_values, na.rm = TRUE)
max_yield <- max(yield_values, na.rm = TRUE)
sd_yield <- sd(yield_values, na.rm = TRUE)

# Print statistics
print(paste("Average Grain Yield (GY):", round(mean_yield, 2)))
print(paste("Minimum Grain Yield (GY):", round(min_yield, 2)))
print(paste("Maximum Grain Yield (GY):", round(max_yield, 2)))
print(paste("Standard Deviation of GY:", round(sd_yield, 2)))

# How many accessions do we have yield data for (non-missing)?
num_yield_obs <- sum(!is.na(yield_values))
print(paste("Number of accessions with GY data:", num_yield_obs))
```

## Visualizing data

We can visualize the distribution of a specific variable using histograms and box plots.

```{r}
# Plotting histogram
hist(barley_pheno_data$Protein, breaks = 50)
# Adding mean
abline(v = mean(barley_pheno_data$Protein), col = "red")
# Adding median
abline(v = median(barley_pheno_data$Protein), col = "blue")
```

```{r}
# Plotting box plot
boxplot(barley_pheno_data$Protein)
```

We can evaluate if there is any general structure that can be identified through PCA.

```{r}
# Scale and run PCA (without our 1st ID column)
pca_res <- prcomp(barley_pheno_data[,-1], scale. = TRUE)

# Summary: variance explained
summary(pca_res)

# PCA scores (samples)
scores <- as.data.frame(pca_res$x)

# PCA loadings (traits contribution)
loadings <- as.data.frame(pca_res$rotation)

# Plot first two PCs
plot(scores$PC1, scores$PC2,
     xlab = paste0("PC1 (", round(summary(pca_res)$importance[2,1]*100, 1), "%)"),
     ylab = paste0("PC2 (", round(summary(pca_res)$importance[2,2]*100, 1), "%)"),
     main = "PCA Scatterplot")
```
