# Module 2.3: General Biostatistics

We will be using an artificial data set of plant traits to practice some general biostatistical analyses. We first need to start by loading the required libraries.

## Libraries

```{r setup, include=FALSE}
library(here)
```

```{r}
# Install required packages if not already installed
packages <- c("tidyverse", "dplyr", "ggplot2", "reshape2",
              "FactoMineR", "factoextra", "corrplot")

installed <- rownames(installed.packages())
for (p in packages) {
  if (!(p %in% installed)) {
    install.packages(p)
  }
}
```

```{r}
# Loading required libraries
library(tidyverse)
library(readxl)
library(dplyr)
library(ggplot2)
library(reshape2)
library(FactoMineR)
library(factoextra)
library(corrplot)
```

## Data import

Lets import our excel data sheet into R.

```{r import-run, include=FALSE}
# Import excel data
data <- read_excel(here("data","PlantData.xlsx"))
```

```{r import, eval=FALSE}
# Import excel data
data <- read_excel("data/PlantData.xlsx")
```

## Exploring data

```{r}
# Lets see how it looks
str(data)
```

We can see that we have a few numeric variables and a few character variables. We want to turn these character variables into factors (categories) for posterior analyses.

```{r}
# Turning into factor
data$Species <- factor(data$Species)
data$Treatment <- factor(data$Treatment)
data$Location <- factor(data$Location)
data$Color <- factor(data$Color)

# Exploring again
str(data)

# Exploring frequencies
table(data$Species)
table(data$Treatment)
table(data$Location)
table(data$Color)
```

## Basic Summaries

Lets perform some basic statistic summaries of our data.

```{r}
summary(data)
```

## Data Distribution

We can use histograms on our numeric variables to explore their distribution.

```{r}
# Histograms
par(mfrow = c(2, 3))

hist(data$Height_cm, main = "Height", xlab = "cm", col = "lightblue", breaks = 10)
hist(data$LeafLength_cm, main = "Leaf Length", xlab = "cm", col = "lightgreen", breaks = 10)
hist(data$LeafWidth_cm, main = "Leaf Width", xlab = "cm", col = "lightgreen", breaks = 10)
hist(data$Biomass_g, main = "Biomass", xlab = "g", col = "orange", breaks = 10)
hist(data$Yield_g, main = "Yield", xlab = "g", col = "purple", breaks = 10)
hist(data$Chlorophyll, main = "Chlorophyll", xlab = "Index", col = "darkgreen", breaks = 10)

par(mfrow = c(1, 1))
```

## Significant differences between two groups

We want to explore the effect of treatment (Control vs Treated) on Height. On our histogram we have seen that height is approximately **symmetric** and close to a **normal distribution,** so we can use a t-test. Even if the distribution is not perfectly normal, we can usually still use a t-test.

### T-test

A **t-test** is a statistical test used to compare **means** between groups (or a group vs. a reference) to see if they are **significantly different**. It assumes data is approximately normally distributed.

```{r}
t.test(Height_cm ~ Location, data = data)
```

We obtain a **p-value** of **0.009387**, which is **less than** our threshold of **0.05**. This means there is a significant difference in heights between our two groups.

**If p-value \< 0.05 → Significant difference between groups**

### Wilcoxon signed-rank test

If our data is not normally distributed (not symmetric or skewed) and we want an alternative to the t-test, we can use a **Wilcoxon signed-rank test**. Instead of comparing means, it compares **medians/distributions** between groups. We now want to explore the effect of treatment (Control vs Treated) on Leaf Length.

```{r}
wilcox.test(LeafLength_cm ~ Treatment, data = data)
```

In this case our **p-value is greater than 0.05**, meaning that there is **no significant difference** between both groups.

## Significant differences between multiple groups

If we wanted to explore the effect of our different Species on Biomass we would use an **Analysis of Variance (ANOVA)**, given we have more than two groups.

**ANOVA is used to test if three or more groups are significantly different!**

### ANOVA

```{r}
# Carrying out ANOVA
a1 <- aov(Biomass_g ~ Species, data = data)

# Printing results
summary(a1)
```

We can see that our obtained **p-value is very small and less than 0.05**, meaning species has a **significant effect** on biomass. We can also make pairwise comparisons between our groups to have a more detailed view of their effect on our experimental measure.

```{r}
# Pairwise comparisons between groups
TukeyHSD(a1)
```

## Significant differences between multiple groups and conditions

If we wanted to evaluate the effect of species AND location on Yield we could use a **Two-way ANOVA**, which allows us to evaluate the effect of each of these conditions and of the interaction between them.

### Two-way ANOVA

```{r}
# Carrying out ANOVA
a2 <- aov(Yield_g ~ Species * Location, data = data)

# Printing results
summary(a2)

# Pairwise comparisons between groups
TukeyHSD(a2)
```

Our results show that on their own, **species and location do have a significant effect** on yield as we have obtained a **p-value smaller than 0.05**: however, the **interaction effect does not seem to have a significant effect** as the obtained p-value is greater than 0.05.

## Pearson Correlation

**Pearson correlation** is a statistic (often written as **r**) that measures the **linear relationship** between two continuous variables.

-   Ranges from **-1 to +1**

    -   **+1** → perfect positive linear relationship (when one increases, the other always increases proportionally)

    -   **0** → no linear relationship.

    -   **-1** → perfect negative linear relationship (when one increases, the other always decreases proportionally).

Now we want to see if there is a relationship between our different numerical values. We can calculate their correlation and plot them as a correlation map.

```{r}
# Lets separate the numeric and factor variables
num_vars <- c("Height_cm","Biomass_g","Yield_g","Chlorophyll",
              "LeafLength_cm","LeafWidth_cm")

# Create a filtered data frame of only our numeric variables
num_df <- data[ , num_vars]
# Calculate our pearson correlation
cor_mat <- cor(num_df, use = "pairwise.complete.obs", method = "pearson")
cor_mat

# Plot results
corrplot(cor_mat, method = "color", type = "upper", addCoef.col = "black",
         tl.col = "black", tl.srt = 45, number.cex = 0.7,
         mar = c(0,0,2,0))
```

## PCA

A **PCA** transforms many correlated variables (Height, Biomass, Yield, Chlorophyll, etc.) into a smaller set of **uncorrelated components** (PC1, PC2, …). Each component aims to capture **as much variance as possible**. It is a common technique used to visualize multi-variate data and identify correlated traits.

```{r}
# Using FactoMineR so we can color by groups easily
pca <- PCA(num_df, scale.unit = TRUE, graph = FALSE)

# Individuals plot colored by Species with 95% ellipses
fviz_pca_ind(
  pca,
  habillage = data$Species, addEllipses = FALSE,
  title = "PCA of Plant Traits (colored by Species)"
)

# Variable loadings
fviz_pca_var(pca, title = "PCA Variable Loadings")
```

-   Direction of arrow = how that variable contributes to the PCA axes (Dim1, Dim2).

-   Length of arrow = strength of contribution (longer = explains more of that axis).

-   Angle between arrows = correlation between variables.

-   Small angle (pointing in the same direction) → strong positive correlation.

-   Opposite direction (180° apart) → strong negative correlation.

-   90° apart → little/no correlation.

## Plots

Lets move on to making some publication-ready plots with `ggplot2`.

### Box plots

Each box shows the **median, quartiles,** and **outliers.**

```{r}
ggplot(data, aes(Treatment, Height_cm, fill = Treatment)) +
  geom_boxplot() +
  labs(title = "Height by Treatment", y = "Height (cm)", x = NULL) +
  theme_minimal() + theme(legend.position = "none")


ggplot(data, aes(Treatment, Biomass_g, fill = Treatment)) +
  geom_boxplot() +
  labs(title = "Biomass by Treatment", y = "Biomass (g)", x = NULL) +
  theme_minimal() + theme(legend.position = "none")
```

```{r}
ggplot(data, aes(Species, Biomass_g, fill = Species)) +
  geom_boxplot() +
  labs(title = "Biomass by Species", y = "Biomass (g)", x = NULL) +
  theme_minimal() + theme(legend.position = "none")

ggplot(data, aes(Species, Chlorophyll, fill = Species)) +
  geom_boxplot() +
  labs(title = "Chlorophyll by Species", y = "Chlorophyll", x = NULL) +
  theme_minimal() + theme(legend.position = "none")
```

```{r}
ggplot(data, aes(Location, Yield_g, fill = Location)) +
  geom_boxplot() +
  labs(title = "Yield by Location (↑ in Greenhouse)", y = "Yield (g)", x = NULL) +
  theme_minimal() + theme(legend.position = "none")

ggplot(data, aes(Location, Chlorophyll, fill = Location)) +
  geom_boxplot() +
  labs(title = "Chlorophyll by Location (↓ in Greenhouse)", y = "Chlorophyll", x = NULL) +
  theme_minimal() + theme(legend.position = "none")
```

### Scatter plot

```{r}
ggplot(data, aes(Height_cm, Biomass_g, color = Species)) +
  geom_point(alpha = 0.8, size = 2.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Height vs Biomass (by Species)",
       x = "Height (cm)", y = "Biomass (g)") +
  theme_minimal()

ggplot(data, aes(Chlorophyll, Yield_g, color = Location)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Yield vs Chlorophyll by Location") +
  theme_minimal()
```

```{r}
ggplot(data, aes(Height_cm, Yield_g, color = Treatment)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(Species ~ Location) +
  labs(title = "Height vs Yield faceted by Species & Location") +
  theme_minimal()
```

### Bar plots

```{r}
# Summarise: mean + standard error
df_summary <- group_by(data, Species, Treatment)
df_summary <- summarise(df_summary, 
                        mean_height = mean(Height_cm, na.rm = TRUE),
                        se_height   = sd(Height_cm, na.rm = TRUE) / sqrt(n()),
                        .groups = "drop")

# Bar plot with error bars
ggplot(df_summary, aes(x = Species, y = mean_height, fill = Treatment)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = mean_height - se_height,
                    ymax = mean_height + se_height),
                position = position_dodge(width = 0.8),
                width = 0.25) +
  labs(title = "Mean Height by Species and Treatment",
       y = "Height (cm)", x = "Species") +
  theme_minimal()
```

### Saving a plot

Many times we will want to save a plot as a `png`, `jpg` or `pdf`. We can easily do this the following way:

```{r, include = FALSE}
# Create plot into an object (in this case p)
p <- ggplot(data, aes(x = Species, y = Biomass_g, fill = Species)) +
  geom_boxplot() +
  theme_minimal()
p

# Save as PNG
ggsave(here("output","boxplot_biomass_species.png"), p, width = 6, height = 4, dpi = 300)

# Or as PDF
ggsave(here("output","boxplot_biomass_species.pdf"), p, width = 6, height = 4)
```

```{r, eval = FALSE}
# Create plot into an object (in this case p)
p <- ggplot(data, aes(x = Species, y = Biomass_g, fill = Species)) +
  geom_boxplot() +
  theme_minimal()
p

# Save as PNG
ggsave("output/boxplot_biomass_species.png", p, width = 6, height = 4, dpi = 300)

# Or as PDF
ggsave("output/boxplot_biomass_species.pdf", p, width = 6, height = 4)
```
