# Module 2.3: General Biostatistics

We will be using a data of wheat plant height values to practice some general biostatistical analyses. We first need to start by importing our data.

```{r setup, include=FALSE}
library(here)
```

```{r import-run, include=FALSE}
# Loading required library
library(readxl)

# Import excel data frame
df <- read_excel(here("data","plantHeights.xlsx"))
```

```{r import, eval=FALSE}
# Loading required library
library(readxl)

# Import excel data frame
df <- read_excel("data/plantHeights.xlsx")
```

Let's look at our data frame.

```{r}
head(df)
str(df)
```

We can see that we have a lot of empty values, we want to remove these before any further analyses.

```{r}
df <- df[complete.cases(df),]
head(df)
```

Our plant height variable is the column `PHT`. We want to visualize the distribution of this variable.

```{r}
# Summary statistics
summary(df$PHT)

# Histogram
hist(df$PHT)

# Box plot
boxplot(df$PHT)
```

We can see that there are two outlier values above 800, which are most likely wrongly annotated. We will remove this for the sake of the analysis.

```{r}
# We want to keep only those values less than 800
df <- df[df$PHT < 800,]

# We plot our histogram aghain
hist(df$PHT, breaks = 50)

# We plot our boxplot again
boxplot(df$PHT)
```

If we look at our data frame, we can see we have year and country information, we will use this to make certain assertions about our data.

```{r}
# Box plot by year
boxplot(df$PHT ~ df$YEAR)

# box plot by country
boxplot(df$PHT ~ df$Country, las = 2, cex.axis = 0.4)
```

As we saw in our histogram, our data is approximately normally distributed, which means we can run a t-test to see if there is a significant difference between the plant heights of two different countries. I will choose AZE and CHN for this case.

```{r}
AZE_PHT <- df[df$Country == "AZE",]
CHN_PHT <- df[df$Country == "CHN",]

# Creating data frame with data only from these two countries
AZE_CHN_PHT <- rbind(AZE_PHT, CHN_PHT)

# Running t-test
t.test(AZE_CHN_PHT$PHT ~ AZE_CHN_PHT$Country)
```

We get a p-value of 0.04, less than our threshold of 0.05. This means we we can reject the null hypothesis of no differentiation between the means of both groups.

**What if we wanted to evaluate more than one variable, not only two countries but all of the countries with more than 100 observations?**

```{r}
# Filter for countries with more than 20 observations
df_filtered <- df[df$Country %in% names(which(table(df$Country) > 100)),]

# Make sure variables are factored
df_filtered$Country <- factor(df_filtered$Country)

# Fit one-way ANOVA
anova_model <- aov(PHT ~ Country, data = df_filtered)

# ANOVA table
summary(anova_model)
```

In this results, the **Sum Sq** shows us the variation in plant height explained by differences between countries, and the variation within countries (which is the one shown in the **Residuals** row). The **F value** is the ratio of between-group variance (countries) to within-group variance (residuals). A large F means country differences are much larger than random variation. Our obtained p-value is extremely small, meaning plant height does significantly differ between countries.

**What if we wanted to know which pairs of countries differed (the individual interactions)?**

```{r}
# We would run the Tukey HSD test
tuk <- TukeyHSD(anova_model)
tuk

# To see the pairs with lower p-values (most significant differences)
# We exctract results and turn into data frame
tuk_df <- as.data.frame(tuk$Country)

# Add pair names
tuk_df$Pair <- rownames(tuk$Country)

# Sort by adjusted p-value
tuk_df_sorted <- tuk_df[order(tuk_df$`p adj`), ]

# Show the pairs with significant differences
tuk_df_sorted[tuk_df_sorted$`p adj` < 0.05,]
```
